---
title: "Informative maps"
format:
  html: default
editor_options: 
  chunk_output_type: console
---

Car fataility data from: The `states` dataset in the `poliscidata` package containing 135 variables on each of the 50 US states.


```{r, echo=FALSE}
#| message: false
#| warning: false

# Initial packages required (we'll be adding more)
library(tidyverse)
library(mdsr)      # package associated with our MDSR book
library(htmltools)
library(glue)
library(leaflet)
library(maps)
library(viridis)
library(sf)
library(rvest)
library(polite)
```


```{r, echo=FALSE}
#| message: false

us_states <- map_data("state")

states <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")  
```


```{r, echo=FALSE}
session <- bow("https://en.wikipedia.org/wiki/List_of_states_and_territories_of_the_United_States_by_population_density", force = TRUE)

result <- scrape(session) |>
  html_nodes(css = "table") |> 
  html_table(header = TRUE, fill = TRUE)
density_table <- result[[1]]

density_data <- density_table |>
  select(1, 2, 4, 5) |>
  filter(!row_number() == 1) |>
  rename(Land_area = `Land area`) |>
  mutate(state_name = str_to_lower(as.character(Location)),
         Density = parse_number(Density),
         Population = parse_number(Population),
         Land_area = parse_number(Land_area)) |>
  select(-Location)
```


```{r, echo=FALSE}
states_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")

states_joined <- states_sf |> 
  mutate(lowercase_name = str_to_lower(name)) |>
  left_join(density_data, by = c("lowercase_name" = "state_name")) |>
  filter(name != "Alaska", name != "Hawaii", name != "Puerto Rico")
```


```{r, echo=FALSE}
#| message: false
#| warning: false

# Get info to draw US states for geom_polygon (connect the lat-long points)
library(maps)
states_polygon <- as_tibble(map_data("state")) |>
  select(region, group, order, lat, long)

# Get info to draw US states for geom_sf and leaflet (simple features object 
#   with multipolygon geometry column)
library(sf)
states_sf <- 
  read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson") |>
  select(name, geometry)

# Load in state-wise data for filling our choropleth maps
#   (Note that I selected my two variables of interest to simplify)
library(poliscidata)   # may have to install first
polisci_data <- as_tibble(poliscidata::states) |>
  select(state, carfatal07, cook_index3)
```

```{r, echo=FALSE}
# Make sure all keys have the same format before joining:
#   all lower case, no internal or external space

states_sf <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")

cleaned_polisci_data <- polisci_data |> 
  mutate(state_same = str_squish(str_to_lower(
    str_replace_all(as.character(state), "(?<=[a-z])(?=[A-Z])", " "))))
```


```{r, echo=FALSE}
# Now we can merge data sets together for the static and the interactive plots

# Merge with states_polygon (static)
polisci_polygon <- cleaned_polisci_data |>
  right_join(states_polygon, by = c("state_same" = "region"))

# Merge with states_sf (static or interactive)

polisci_sf <- states_sf |>
  mutate(name_lower = str_squish(str_to_lower(name))) |>
  left_join(cleaned_polisci_data, by = c("name_lower" = "state_same")) |>
  filter(!name %in% c("Alaska", "Hawaii", "Puerto Rico"))
```

Numeric variable (static plot):

```{r, echo=FALSE}
polisci_polygon |>
  filter(!is.na(state_same)) |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = carfatal07), color = "yellow", linewidth = 0.2) +
  coord_map() +
  theme_void() +
  scale_fill_viridis(name = "Car Fatalities per 100,000 in 2007") +
  labs(title = "Car Fatalities by State in 2007")
```

**Car fatality rates tend to be higher in south-eastern states more than anywhere else in the United States, especially in Mississippi, which has the highest rate. In contrast, states like Minnesota have a lot lower fatality rates.**

Numeric variable (interactive plot):

```{r, echo=FALSE}
pal_car <- colorNumeric(
  palette = "Greens",
  domain = polisci_sf$carfatal07,
  na.color = "grey")

polisci_sf <- polisci_sf |>
  mutate(labels = str_c(name, ": ", round(carfatal07, 1), " fatalities per 100k"))

labels_car <- lapply(polisci_sf$labels, HTML)

leaflet(polisci_sf) |>
  setView(-96, 37.8, 4) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal_car(carfatal07),
    weight = 1,
    opacity = 1,
    color = "seagreen",
    fillOpacity = 0.7,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#555",
      bringToFront = TRUE),
    label = labels_car,
    labelOptions = labelOptions(
      textsize = "15px",
      direction = "auto"))
```

**In Mississippi there are about 31.6 car fatalities per 100,000 people. This is higher than in any other state in the United States. This is 19.9 more fatalities per 100,000 people than in Minnesota, which has a fatality car accident rate of 11.7 per 100,000. **

Categorical variable (static plot):

```{r, echo=FALSE}
# be really careful with matching color order to factor level order

polisci_polygon |>
  filter(!is.na(cook_index3)) |>
  ggplot(aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = cook_index3), color = "white", linewidth = 0.2) +
  coord_map() +
  theme_void() +
  scale_fill_manual(
    values = c("More Rep" = "firebrick", "Even" = "purple", "More Dem" = "steelblue"),
    name = "Cook Index") +
  labs(title = "Political view by State")

```

**States along the East Coast tend to lean more Democratic, particularly in the Northeast, while many states in the middle of the country lean to the Republican side. There is also a Republican states band through the middle of the United States in the midwest and south, while Democratic-leaning states are more concentrated on the coasts. **

Categorical variable (interactive plot):

```{r, echo=FALSE}
pal_cook <- colorFactor(
  palette = c("firebrick", "purple", "steelblue"),
  levels = c("More Rep", "Even", "More Dem"),
  na.color = "grey")

polisci_sf <- polisci_sf |>
  mutate(labels = str_c(name, ": ", as.character(cook_index3)))

labels_cook <- lapply(polisci_sf$labels, HTML)

leaflet(polisci_sf) |>
  setView(-96, 37.8, 4) |>
  addTiles() |>
  addPolygons(
    fillColor = ~pal_cook(cook_index3),
    weight = 1,
    opacity = 1,
    color = "seagreen",
    fillOpacity = 0.75,
    highlightOptions = highlightOptions(
      weight = 3,
      color = "#555",
      bringToFront = TRUE),
    label = labels_cook,
    labelOptions = labelOptions(
      textsize = "15px",
      direction = "auto")) |>
  addLegend(pal = pal_cook, values = ~cook_index3, opacity = 0.7,
            title = "Cook Index",
            position = "bottomright")
```

**The interactive map presents the same categorical information as the static version but it allows you to explore individual states more closely. When you hover over the states you can identify the political leaning easily. The map also also shows that "Even" states are more scattered rather than concentrated in a region. **
